Estimating duration in the presence of misclassification

### 2023-12-21 ATACCC.tex
easy edits

\Cref{E-inc-prev} showed that the duration of PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
The aim of this chapter is to estimate the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
I model a biomarker, the log viral load, over the course of an infection.\todo{make sure log viral load and Ct values are introduced in chapter 2}
From this model, I derive the distribution of the duration of PCR positivity.

@@

The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
The viral load model produces a deterministic trajectory for the viral load of an individual, a function of the model parameters.
The observation model then describes how the PCR testing process produces noisy observations of this viral load.
The observation model includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
\todo{Make sure limit of detection is introduced previously when I first talk about PCR}
% The limit of detection means that viral loads below a certain threshold are not detected by the PCR test.
% They can be represented as a censoring of the viral load at the limit of detection.

@@

PCR tests are noisy observations of the viral load.
A positive PCR test has a Ct value.
Ct values are linearly related to viral load (see\todo{make sure relationship between Ct values and viral load is introduced previously}).
A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
Positive PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.

>>>>

\Cref{E-inc-prev} showed that the duration of PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
The aim of this chapter is to estimate the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
I model a biomarker, the log viral load, over the course of an infection.
From this model, I derive the distribution of the duration of PCR positivity.

@@

The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
The viral load model produces a deterministic trajectory for the viral load of an individual, a function of the model parameters.
The observation model then describes how the PCR testing process produces noisy observations of this viral load.
The observation model includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
% The limit of detection means that viral loads below a certain threshold are not detected by the PCR test.
% They can be represented as a censoring of the viral load at the limit of detection.

@@

PCR tests are noisy observations of the viral load.
A positive PCR test has a Ct value.
Ct values are linearly related to viral load (see \cref{E-biology-data:sec:PCR})
A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
Positive PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.

### 2023-12-21 cis-imperfect-testing.tex
easy edits

Ideally, the test sensitivity would be estimated from the data.
However, this would require a model for time-varying test sensitivity to be incorporated into the model.
Furthermore, it is unclear that such a model would be identifiable because it would not be clear whether to attribute negative results to false negatives or to recoveries\todo{cite something on this}.

>>>>

Ideally, the test sensitivity would be estimated from the data.
However, this would require a model for time-varying test sensitivity to be incorporated into the model.
Furthermore, it is unclear that such a model would be identifiable because it would not be clear whether to attribute negative results to false negatives or to recoveries\todo{cite identifiability issues with varying sensitivity}.

### 2023-12-21 incidence-prevalence.tex
easy edits

Afterwards, in \cref{SEIR}, I introduce \emph{mechanistic models}.
Mechanistic models explicitly describe the population and transmission between individuals.
They vary greatly in their details and realism~\autocite{murilloMultiscale}.
\todo[inline]{Make sure this section is coherent: a bunch of stuff from the SEIR chapter has been copied and pasted in}

@@

The other defining property of a Poisson process is that it has \emph{independent increments}.
A counting process has independent increments if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
\todo{check if this should be $<$ or $\leq$}
In the context of a Poisson process, the independent increment property implies that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
Hence, $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.

@@

Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters\todo{REF stochastic models}.
Stochastic models are most commonly specified in discrete time, using a Markov assumption.
A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).

@@

The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
The name SIR derives from the structure of the model.
In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: susceptible, infectious, or recovered.
% The state of the system is the proportion of the population in each compartment.
In the SIR model, individuals are assumed to be homogeneous except for their disease state, \ie they are otherwise identical.

@@

The SIR model assumes that individuals are immediately infectious upon infection.
This is unrealistic for most diseases, as there is a period between an individual being infected and them being infectious, known as the \emph{latent period}.
\todo{I will probably end up defining the latent period in the intro, so should just ref that}
Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.

The SIR model can be extended to include a latent period by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.

@@

MCMC using a Normal proposal distribution can be inefficient.
In particular, it often fails to propose moves into the tails of the posterior distribution\todo{any suggested refs for this (Paul?) - we've discussed this lots but I've never read it anywhere}.
\emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
HMC simulates a physical system travelling through the posterior space with momentum.
The momentum increases in the direction the posterior density increases, using the gradient.
\todo{add something about leapfrog steps}
This means that the sampler tends to propose moves with higher posterior density.
The Metropolis-Hastings algorithm is then used to accept or reject the proposed move.

The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the number of leapfrog steps to take in the Hamiltonian dynamics.
NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
Stan implements \emph{automatic differentiation}\todo{ref automatic diff} to calculate the gradient of the posterior density without any user input.
Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known.
For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.

>>>>

Afterwards, in \cref{SEIR}, I introduce \emph{mechanistic models}.
Mechanistic models explicitly describe the population and transmission between individuals.
They vary greatly in their details and realism~\autocite{murilloMultiscale}.

@@

The other defining property of a Poisson process is that it has \emph{independent increments}.
A counting process has independent increments if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
In the context of a Poisson process, the independent increment property implies that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
Hence, $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.

@@

Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters\todo{Cite stochastic models}.
Stochastic models are most commonly specified in discrete time, using a Markov assumption.
A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).

@@

The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{cite SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
The name SIR derives from the structure of the model.
In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: susceptible, infectious, or recovered.
% The state of the system is the proportion of the population in each compartment.
In the SIR model, individuals are assumed to be homogeneous except for their disease state, \ie they are otherwise identical.

@@

The SIR model assumes that individuals are immediately infectious upon infection, neglecting the latent period (see \cref{E-biology-data:sec:natural-history}).
Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
The SIR model can be extended by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.

@@

MCMC using a Normal proposal distribution can be inefficient.
In particular, it often poorly explores the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
\emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
HMC simulates a physical system travelling through the posterior space with momentum.
The momentum increases in the direction the posterior density increases, using the gradient.
\todo{add something about leapfrog steps}
This means that the sampler tends to propose moves with higher posterior density.
The Metropolis-Hastings algorithm is then used to accept or reject the proposed move.

The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the number of leapfrog steps to take in the Hamiltonian dynamics.
NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
Stan implements \emph{automatic differentiation}\todo{cite automatic diff} to calculate the gradient of the posterior density without any user input.
Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known.
For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.

### 2023-12-21 transmission.tex
Fix references
easy edits

The majority of the confidence intervals contain the nominal probability (see \cref{SEIR:fig:sim-coverage}).
There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow.
It is well-known that random walk MCMC, as used here, can poorly explore the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
\todo{cite issues with random-walk MCMC exploring space}
This is similar when considering coverage of the derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
\begin{figure}
    \includegraphics{SEIR/sim/coverage}
    \caption[Coverage of simulation study]{%
        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter; see text for full definition.
        Vertical lines show the nominal coverage.
    }
    \label{SEIR:fig:sim-coverage}
\end{figure}
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
    \caption[Coverage of simulation study (derived quantities)]{%
        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for incidence (top) and prevalence (bottom); see text for full definition.
        Horizontal lines show the nominal coverage.
    }
    \label{SEIR:fig:sim-inc-prev}
\end{figure}

@@

The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
Incidence is smooth within each week but shows discontinuities between weeks.
This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
A constant $\matr{\beta}(t)$ leads to exponential growth over short time periods, once the initial dynamics have died out.
When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
For Yorkshire, the peak is estimated to occur in late October.
For all other regions, it is estimated to occur in late December or Early January, the period during which severe social distancing measures were applied.
\begin{figure}
    \includegraphics{SEIR/CIS/incidence}
    \caption[Posterior estimates of incidence]{%
        Posterior estimates (95\% CrI) of incidence by region and age.
        See main text for discussion.
    }
    \label{SEIR:fig:incidence}
\end{figure}

% \begin{figure}
%     \includegraphics{SEIR/CIS/p_peak}
%     \caption[Posterior estimates of peak incidence timing]{%
%         Posterior probability of the peak incidence for each region and age combination occurring on each day.
%     }
%     \label{SEIR:fig:peak-incidence}
% \end{figure}

>>>>

The majority of the confidence intervals contain the nominal probability (see \cref{SEIR:fig:sim-coverage}).
There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow.
This is a common issue with random walk MCMC (see \cref{E-inc-prev:sec:MCMC}).
This is similar when considering coverage of the derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
\begin{figure}
    \includegraphics{SEIR/sim/coverage}
    \caption[Coverage of simulation study]{%
        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter; see text for full definition.
        Vertical lines show the nominal coverage.
    }
    \label{SEIR:fig:sim-coverage}
\end{figure}
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
    \caption[Coverage of simulation study (derived quantities)]{%
        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for incidence (top) and prevalence (bottom); see text for full definition.
        Horizontal lines show the nominal coverage.
    }
    \label{SEIR:fig:sim-inc-prev}
\end{figure}

@@

The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
Incidence is smooth within each week but shows discontinuities between weeks.
This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
A constant $\matr{\beta}(t)$ leads to exponential growth over short time periods, once the initial dynamics have died out.
When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
For Yorkshire, the peak is estimated to occur in late October.
For all other regions, it is estimated to occur in late December or Early January, the period during which severe social distancing measures were applied.
\begin{figure}
    \includegraphics{SEIR/CIS/incidence}
    \caption[Posterior estimates of incidence]{%
        Posterior estimates (95\% CrI) of incidence by region and age.
        See main text for discussion.
    }
    \label{SEIR:fig:incidence}
\end{figure}
\begin{figure}
    \includegraphics{SEIR/CIS/p_peak}
    \caption[Posterior estimates of peak incidence timing]{%
        Posterior probability of the peak incidence for each region and age combination occurring on each day.
    }
    \label{SEIR:fig:peak-incidence}
\end{figure}
